Using SIMD and SIMT vectorization to evaluate sparse chemical kinetic Jacobian matrices and thermochemical source terms

### 2018-08-12 paper.tex
dr. sung edits

In this article we
\begin{itemize}
 \item Derive and \add{reproducibly} verify a new Jacobian formulation that greatly increases sparsity\delete{: \texttt{pyJac v2}};
 \item \delete{Enable}\add{Detail the implementation of} cross-platform SIMD\slash SIMT vectorization for CPUs, GPUs, and other accelerators;
 \item \add{Investigate the performance of SIMD\slash SIMT-vectorization for a wide of range chemical kinetic models, and compare with the previous version of our analytical chemical kinetic Jacobian code~\cite{Niemeyer:2016aa}; and finally}
 \item Discuss future extensions to this work as well as several promising directions for SIMD\slash SIMT vectorization in reactive-flow simulations.
\end{itemize}
This work builds upon our previous analytical chemical kinetic Jacobian code, \texttt{pyJac}~\cite{Niemeyer:2016aa}\add{, using the new formulation, \texttt{pyJac v2}, to achieve these goals}.
To our knowledge is the first open-source, verified effort that vectorizes the evaluation of chemical-kinetic source terms and Jacobian matrices for any chemical model on a wide selection of platforms.

@@

In addition, \texttt{loo.py} allows the user to more easily make changes to the structure of the generated program, e.g., the data ordering, vectorization, and threading patterns, as well as switch the target language for code generation (and more simply extend to additional languages, e.g., CUDA).
Further, \texttt{loo.py} can execute developed subroutines from Python (natively for C code, or via \texttt{PyOpenCL}~\cite{kloeckner_pycuda_2012} for OpenCL), enabling unit testing\slash verification for each component of the Jacobian or source terms; \add{the unit testing suite also helps ensure that bugs are not present in less commonly used code-paths}.
The source terms and sub-components thereof (e.g., rates of progress, pressure-modification terms) are directly compared with Cantera~\cite{Cantera}, while the automatic differentiation code Adept~\cite{hogan2014fast,adept-v11} provides reference values for Jacobian sub-components.
We use the Portable OpenCL (POCL) implementation~\cite{poclIJPP} and OpenMP~\cite{dagum1998openmp} to perform OpenCL and C unit testing, respectively, on the continuous-integration framework Travis CI~\cite{travis:2018}.
We will discuss verification of the complete (as opposed to the sub-component testing discussed here) generated source-terms and Jacobian codes in detail in~\cref{s:verification}.

@@

To demonstrate the type of issue discussed, we created a minimum working example~\cite{Nvidia_mwe} that shows a failure of Nvidia's OpenCL runtime corresponding to the GPU driver version \num{375.26} on a Tesla \gpunew/ GPU; simply changing to another runtime (e.g., Intel) produces the correct result---with no code changes or recompilation.
This provides a particularly vexing problem for the programmer, since little can often be done to resolve the issue; thankfully, in this case we were able to upgrade the driver version to resolve the problem.
Further, as noted throughout this work, certain code-generation patterns can break OpenCL execution\slash vectorization, e.g., the failure of POCL to achieve vectorization or Intel OpenCL's failure to vectorize the finite-difference Jacobian.
Indeed the vectorization process attempted by most OpenCL runtimes (and thus, reasons for incorrect\slash unvectorized code output) is obscured from the user, making reasoning about errors or performance trends quite difficult.
Thankfully, \texttt{loo.py} allows relatively easy switching between output languages; the most significant code change requires building of the wrapper that initializes\slash transfers memory and calls the source-rate\slash Jacobian kernel.

\add{These issues make it critical to provide adequate implementation details (e.g., runtime version, platform, etc., as given in~\cref{s:test_platforms}) for codes utilizing OpenCL in order to enable reproducible results.}

@@

In this work, we developed automatically generated OpenCL codes for SIMD and SIMT-vectorized evaluations of thermochemical source-term and sparse\slash dense chemical kinetic Jacobian matrices.
The main contributions of this work are:
\begin{itemize}
 \item Deriving and \add{reproducibly} verifying a new Jacobian formulation \delete{for \texttt{pyJac}} that greatly increases sparsity;
 \item \delete{Extending \texttt{pyJac} to enable}\add{Enabling} vectorized execution on the CPU, GPU, and other accelerators; and
 \item Achieving significant speedups over a strictly parallel Jacobian and source-term evaluation on SIMD-enabled processors (e.g., CPUs).
\end{itemize}

>>>>

In this article we
\begin{itemize}
 \item Derive and verify a new Jacobian formulation that greatly increases sparsity\delete{: \texttt{pyJac v2}};
 \item \delete{Enable}\add{Detail the implementation of} cross-platform SIMD\slash SIMT vectorization for CPUs, GPUs, and other accelerators;
 \item \add{Investigate the performance of SIMD\slash SIMT-vectorization for a wide range of chemical kinetic models, and compare with the previous version of our analytical chemical kinetic Jacobian code~\cite{Niemeyer:2016aa}; and finally}
 \item Discuss future extensions to this work as well as several promising directions for SIMD\slash SIMT vectorization in reactive-flow simulations.
\end{itemize}
This work builds upon our previous analytical chemical kinetic Jacobian code, \texttt{pyJac}~\cite{Niemeyer:2016aa}\add{, using the new formulation, \texttt{pyJac v2}, to achieve these goals}.
To our knowledge is the first open-source, verified effort that vectorizes the evaluation of chemical-kinetic source terms and Jacobian matrices for any chemical model on a wide selection of platforms.

@@

In addition, \texttt{loo.py} allows the user to more easily make changes to the structure of the generated program, e.g., the data ordering, vectorization, and threading patterns, as well as switch the target language for code generation (and more simply extend to additional languages, e.g., CUDA).
Further, \texttt{loo.py} can execute developed subroutines from Python (natively for C code, or via \texttt{PyOpenCL}~\cite{kloeckner_pycuda_2012} for OpenCL), enabling unit testing\slash verification for each component of the Jacobian or source terms; \add{the unit testing suite also helps ensure that bugs are not present in less commonly used code-paths, or are introduced by future code changes}.
The source terms and sub-components thereof (e.g., rates of progress, pressure-modification terms) are directly compared with Cantera~\cite{Cantera}, while the automatic differentiation code Adept~\cite{hogan2014fast,adept-v11} provides reference values for Jacobian sub-components.
We use the Portable OpenCL (POCL) implementation~\cite{poclIJPP} and OpenMP~\cite{dagum1998openmp} to perform OpenCL and C unit testing, respectively, on the continuous-integration framework Travis CI~\cite{travis:2018}.
We will discuss verification of the complete (as opposed to the sub-component testing discussed here) generated source-terms and Jacobian codes in detail in~\cref{s:verification}.

@@

To demonstrate the type of issue discussed, we created a minimum working example~\cite{Nvidia_mwe} that shows a failure of Nvidia's OpenCL runtime corresponding to the GPU driver version \num{375.26} on a Tesla \gpunew/ GPU; simply changing to another runtime (e.g., Intel) produces the correct result---with no code changes or recompilation.
This provides a particularly vexing problem for the programmer, since little can often be done to resolve the issue; thankfully, in this case we were able to upgrade the driver version to resolve the problem.
Further, as noted throughout this work, certain code-generation patterns can break OpenCL execution\slash vectorization, e.g., the failure of POCL to achieve vectorization or Intel OpenCL's failure to vectorize the finite-difference Jacobian.
Indeed the vectorization process attempted by most OpenCL runtimes (and thus, reasons for incorrect\slash unvectorized code output) is obscured from the user, making reasoning about errors or performance trends quite difficult.
Thankfully, \texttt{loo.py} allows relatively easy switching between output languages; the most significant code change requires building of the wrapper that initializes\slash transfers memory and calls the source-rate\slash Jacobian kernel.
\add{These issues make it critical to provide adequate implementation details (e.g., runtime version, platform, etc., as given in~\cref{s:test_platforms}) for codes utilizing OpenCL in order to enable reproducible results.}

@@

In this work, we developed automatically generated OpenCL codes for SIMD and SIMT-vectorized evaluations of thermochemical source-term and sparse\slash dense chemical kinetic Jacobian matrices.
The main contributions of this work are:
\begin{itemize}
 \item Deriving and verifying a new Jacobian formulation \delete{for \texttt{pyJac}} that greatly increases sparsity;
 \item \delete{Extending \texttt{pyJac} to enable}\add{Enabling} vectorized execution on the CPU, GPU, and other accelerators; and
 \item Achieving significant speedups over a strictly parallel Jacobian and source-term evaluation on SIMD-enabled processors (e.g., CPUs).
\end{itemize}
